Management induced changes of soil organic carbon on global croplands

### 2020-12-11 SOCbudgetPaper/SOCbudget.tex
compliling version

%% 2-column papers and discussion papers
\RequirePackage[2020/02/02]{latexrelease}
\documentclass[gc, manuscript]{copernicus}

@@

\Author[1]{Kristine}{Karstens}
\Author[1]{Benjamin Leon}{Bodirsky}
\Author[2]{Your}{Name}
\Author[1]{Alexander}{Popp}

\affil[1]{Potsdam-Institut of Climate Impacts Research, Potsdam, Germany}
\affil[2]{Your affiliation}

@@

\begin{abstract}
Soil organic carbon (SOC) is one of larges carbon stocks on Earth. It is three times larger than the vegetation pool and more than twice as big as the atmospheric pool, when looking into the first meter of the earth soil profile. Human cropping activities led and still lead to a depletion of SOC though, which are so far not well represented in global assessments of historic carbon emissions. While SOC models often represent well the biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated on the global scale. Here we create a spatial explicit data set for agricultural management on cropland, especially for crop residue and manure management, based on global historic production (FAOSTAT), and land-use (LUH2) data; and combine it with the IPCC Tier 2 approach to create a half degree resolution SOC stock changes on mineral soils. We estimate that due to arable farming soils have lost over 37 GtOC of which 4 GtOC have been regained within the period 1975--2010. We show that, our results on global scale based on Tier 2 IPCC methodology are in good agreement with Tier 1 default assumptions. We also find that SOC is very sensitive to management decision such as residue recycling indicating the necessity to incorporated better management data in soil models.
\end{abstract}

@@

Soil Organic Carbon (SOC), the amount of organic carbon stored through the earth's soil, is the largest terrestrial carbon pool, exceeding the carbon in the atmospheric and vegetation pools multiple times (Batjes). As such, even small changes in drivers of SOC may thus lead to substantial shifts in earth carbon cycle and influence the amount of CO2 in the atmosphere (ref. permafrost melting). The specific amount of carbon stored in the soil is, however, uncertain, with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile (Batjes, 1996; more). The quality of SOC maps has markedly improved in recent decades, along with better understanding of the factors driving the global magnitude, distribution, and dynamics of SOC pools. Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles to determine SOC variations over space and time.
Human intervention, including land cover change and land management, has however added a further driver to SOC change, which alters terrestrial carbon pools over much shorter time scales and is likely one of the most dominant driver of SOC changes on managed land today (cite). Recent studies identify the anthropogenic SOC debt for the first meter of the soil profile at around 116 GtC (Sanderman et al.), compared to previous estimates between 60-130 GtC (Lal, 2006). Other studies have focused more closely on spatially disaggregation of SOC changes via advanced digital soil mapping techniques (S-World; Stoorvogel 2, 2017) or better representation of biogeochemical processes within SOC dynamics (Hararuk, 20XX). While providing a rough estimate of the order of magnitude of change, these studies lack a detailed consideration of land management, especially of agricultural activities.
Field-scale models (cite Daycent, RothC, Ecosse, C-Tool) are able to capture these land management impacts by using detailed information on crop yield levels, fertilizer inputs and various other on-farm measures. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global extent remains a complex challenge. Our study combines a spatially explicit estimate of agricultural management data on the global level with a 3-pool SOC model parametrized for global croplands. This allows us to estimate SOC stock change factors, as well as organic carbon flow dynamics within the agricultural system. We thereby consider change in SOC caused by historical land cover change as well as of different agricultural management practices, including residue recycling, manure amendments, irrigation, and tillage. We provide the first global, spatially explicit SOC loss maps that considers detailed agricultural management.
This paper will first introduce in the methods section the basic concept of SOC dynamics as applied in this study. Additionally, we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines. Lastly, we continue with a detailed description of the global gridded management data used here, including crop production levels, residue recycling rates, manure amendments, and the adoption of irrigation and tillage practices. In our results section we focus on the SOC dynamics of global croplands by (1) analysing the spatially explicit distribution and depletion of SOC from 1975 to 2010, (2) comparing impacts of different management effects on SOC emissions and (3) quantify global agricultural carbon flows and stocks to compare the importance of various management aspects. Finally, we discuss and compare our findings -- including their implications for SOC model development -- and conclude with an outlook on the ability of SOC management to mitigate climate change and contribute to negative greenhouse gas emissions.
\newpage

\hypertarget{method}{%
\section{Method}\label{method}}

We compile calculations as open-source R packages available at github.com/pik-piam/mrcommons (management related functions) and github.com/pik-piam/mrSOCbudget (soil dynamic related functions), which are both based on the MADRaT package (package citation?), a framework which aims to improve reproducibility and transparency in data processing.
In the following section we outline the most important relationships and assumptions. Table \ref{append:subsection2mrfunctions} provides further information on corresponding code within the R packages.

\hypertarget{sec:carbonbudget}{%
\subsection{SOC Stocks and Stock changes following the Tier 2 steady-state method}\label{sec:carbonbudget}}

Following the tier 2 approach of the refinement of IPCC guidelines vol.~4 (\citet{ipcc_2019_2019}; short Tier 2 steady-state method), we estimate soil organic carbon (SOC) stocks and their change over time for cropland on half-degree resolution from 1975 to 2010. We assume the current SOC state converges towards a steady state, which itself is depending on biophysical, climatic and agronomic conditions. Therefore, we conduct the following three steps within each yearly time-step:
(1) We calculate annual land-use (sub-)type-specific steady states and decay rates for SOC stocks,
(2) we account for land conversion by transferring SOC from and to another land-use type represented as natural vegetation,
(3) we estimate SOC stocks and changes based on the stocks of the previous time period, the steady state stocks and the decay rate.

\hypertarget{steady-state-soc-stocks-and-decay-rates}{%
\subsubsection{Steady-state SOC stocks and decay rates}\label{steady-state-soc-stocks-and-decay-rates}}

In a simple first order kinetic approach the steady-state soil organic carbon stocks \(SOC^{eq}\) are given by
\begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k} \qquad\forall\quad i,t
\label{eq:inoutflow}
\end{equation}
with \(C^{\textrm{in}}\) being carbon inputs to the soil, \(k\) denoting the soil organic carbon decay rate; as well as \(i\) representing grid cell identifier and \(t\) years. We use for our calculations the Tier 2 steady-state method, which assumes three soil carbon sub-pools (active, slow and passive) and interactions between them, following the approach in the Century model (Parton et al, 1988). Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are the key components to determining steady-state SOC stocks.

\textbf{Carbon Inputs to the Soil}

We account for different carbon input sources depending on the two land-use types we distinguish: croplands and natural vegetated land as representative for all other land use (see table \ref{tab:datasourceinputs}). Carbon sources for cropland are recycled crop residues, below ground biomass of crops (for both see \ref{sec:residues}) and recycled manure (see \ref{sec:livstmanure}); for natural vegetation litterfall (including fine root turnover) \citep{schaphoff_lpjml4_2018} is the only source of carbon inflow to the soil.

Following the IPCC carbon accounting methodology, carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content (see \citet{ipcc_2019_2019}). For each component the sum of all carbon input sources is allocated to the respective SOC sub-pools via transfer coefficients. This implies that both the amount of carbon and its structural composition determine the effective inflow. Data sources for all considered carbon inputs as well as for lignin and nitrogen content can be found in table \ref{tab:datasourceinputs}.

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types }
 \begin{tabular}{l l l l}
 \tophline
  \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
 \middlehline
 \multirow{3}{*}{Cropland} & residues & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2]  \\
                            & dead below ground biomass of crops & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2] \\
                            & manure & FAOSTAT, Isabelle [2, sec:manure] & default values given by [2] \\
                            \hline
  Natural vegetation        & annual litterfall & LPJmL4 [4]& \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut Nitrogen and lignin content of tree compartments used in CENTURY [4] \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}

\textbf{SOC decay}

The sub-pool specific decay rates are influenced by climatic conditions, biophysical and biochemical soil properties as well as management factors that all vary over time (t) and space (i). Following the Tier 2 steady-state method we consider temperature (temp), water (wat), sand fraction (sf) and tillage (till) effects to account for spatial variation of decay rates. Thus, \(k_{sub}\) is given by

\begin{equation}
\begin{aligned}
& k_{active,t,i}  & = &~ k_{active}  ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ &\cdot~ till_{t,i} ~ & \cdot~ sf_{t,i}\\
& k_{slow,t,i}    & = &~ k_{slow}    ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ &\cdot~ till_{t,i} ~ &\\
& k_{passive,t,i} & = &~ k_{passive} ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ & ~ &
\label{eq:decayrates}
\end{aligned}
\end{equation}

For cropland, we distinguish the effect of different tillage (see \ref{sec:tillage}) and irrigation (see \ref{sec:irrigation}) practices on decay rates, whereas for natural vegetation, we assume rainfed and non-tilled conditions. Data sources as well as drivers considered for each land-use types are shown in table \ref{tab:datasourcedecay}. To account for variations of decay rates within each grid cell due to different tillage and irrigation regimes, average rates based on area shares are calculated.

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types}
 \begin{tabular}{l l l l}
 \tophline
  land-use types   & type of decay driver & parameter use to represent driver & data source \\
 \middlehline
 \multirow{2}{*}{all} & Soil quality & Sand fraction of the first 0-30 cm &  [SoilGrids]  \\
                      \cline{2-4}

                      & Mircobial activity & air temperature & [CRUp4.0] \\
                      \cline{2-4}

                      & Water restriction & precipitation \& potential evapotranspiration & [CRUp4.0] \\
                      \cline{1-4}
\multirow{2}{*}{\begin{minipage}[t]{0.2\columnwidth}\raggedright\strut Cropland\\(additionally)\strut\end{minipage}} & Water restriction*  & irrigation  & [sec:irrigation] \\ 
                      \cline{2-4}

                      & Soil disturbance & tillage & [sec:tillage] \\
 \bottomhline
 \end{tabular}
 \belowtable{}
 \label{tab:datasourcedecay}
 \end{table*}

\hypertarget{soc-transfer-between-land-use-types}{%
\subsubsection{SOC transfer between land-use types}\label{soc-transfer-between-land-use-types}}

We calculate SOC stocks based on the area shares of land-use types (lut) within the half-degree grid cells (i). If land is converted from one land-use type into others (!lut), the respective share of the SOC stocks is reallocated. We account for land conversion at the beginning of each time step \(t\) by calculating a preliminary stock \(SOC_{lut,t*}\) via

\begin{equation}
SOC_{lut,t*} = SOC_{lut,t-1} - \frac{SOC_{lut,t-1}}{A_{lut,t-1}} \cdot  AR_{lut,t} + \frac{SOC_{!lut,t-1}}{A_{!lut,t-1}} \cdot  AE_{lut,t} \qquad \forall\quad sub, i  
\label{eq:ctransfer}
\end{equation}

with \(A\) being the area, \(AR\) the area reduction and \(AE\) the area expansion for a given land-use type \(lut\). Note that \(!lut\) denotes the sum over all other land-use types, which decreases in the specific time step \(t\). Data sources and methodology on land-use states and changes are described in \ref{sec:landuse}.

\hypertarget{total-soc-stocks-and-stock-changes}{%
\subsubsection{Total SOC stocks and stock changes}\label{total-soc-stocks-and-stock-changes}}

Carbon stocks \(SOC\) for each sub-pool (sub) converge towards the calculated steady-state stock \(SOC^{eq}\) for each land-use type (lut), each sub-pool (sub) and each annual time step (t) like

\begin{equation}
SOC_{t} = SOC_{t-1} + (SOC^{eq}_{t} - SOC_{t-1}) \cdot k_{t} \cdot 1\unit{a} \qquad \forall\quad\quad lut, sub, i.
\label{eq:SOCstate}
\end{equation}

Reformulating this equation, we obtain a mass balance equation as follows

\begin{equation}
SOC_{t} = SOC_{t-1} - \underbrace{SOC_{t-1} \cdot k_{t} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{eq}_{t} \cdot k_{t} \cdot 1\unit{a}}^{\text{input (using equation (1))}}  \qquad \forall\quad lut, sub, i.
\label{eq:steadystate2budget}
\end{equation}

The global SOC stock for each time step can then be calculated via

\begin{equation}
SOC_{t} = \sum_{i} \underbrace{\sum_{lut} \overbrace{\sum_{sub} SOC_{lut, sub, t, i}.}^{\text{$SOC_{lut, t, i}$ -- land-use type specific SOC stock within cell}}}_{\text{$SOC_{t, i}$ -- total SOC stock within cell}}
\label{eq:totalstock}
\end{equation}

According to the IPCC guidelines SOC changes can be expressed as the difference of two consecutive years (see Eq. 5.0A in \citep{ipcc_2019_2019}). This however, will also include naturally occurring changes due to climatic variance over time. For our study we defined the absolute and relative SOC changes in relation to a potential natural state \(SOC_{pnv}\) under the same climatic conditions at time \(t\) in grid cell \(i\), that is based on the natural vegetation SOC calculations as defined above without accounting for land conversion from cropland at any time. The absolute changes \(\Delta SOC\) and relative changes \(F^{SCF}\) a thus given by

\begin{equation}
\Delta SOC_{t, i} = SOC_{t, i} - SOC_{pnv, t, i}\quad, \qquad \quad  F^{SCF}_{t, i} = \frac{SOC_{t, i}}{SOC_{pnv, t, i}}.
\label{eq:stockdiff}
\end{equation}

Note that the absolution changes \(\Delta SOC\) can be also interpreted as the SOC gap due to human cropping activities; whereas relative changes \(F^{SCF}\) can be denoted as stock change factors as defined within the IPCC guidelines of 2006 \citep{ipcc_2006_2006}. Note that the SOC gap is equivalent to the negated cumulative SOC emission (\citep{pugh_simulated_2015}) also referred to as SOC debt (\citep{sanderman_soil_2017}) of human land use.

\hypertarget{initialization-of-soc-pools}{%
\subsubsection{Initialization of SOC pools}\label{initialization-of-soc-pools}}

To initialize all SOC sub-pools we assume that cropped land as well as natural vegetation are in a steady state for the specific configuration present within the initialization year 1961. We assume after a spin-up period of 15 years reliable results start from 1975 and improve over time, since dependency on the initial conditions decreases.

\hypertarget{sec:tier1}{%
\subsection{SOC Stocks and Stock changes following Tier 1}\label{sec:tier1}}

Additionally, to the tier 2 approach of the refinement of IPCC guidelines vol.~4 \citep{ipcc_2019_2019} and the detailed analysis of management data coming with it, SOC changes can be estimated using the IPCC tier 1 approach of IPCC guidelines vol.~4 (\citet{ipcc_2006_2006}, \citet{ipcc_2019_2019}). Here, stocks are calculated via stock change factors (\(F^{SCF}\)) given by the IPCC for the topsoil (0-30 cm) and based on observational data. \(F^{SCF}\) are differentiated by different crop, management and input systems (here summarized under \(m\)) reflecting different dynamics under changed in- and outflows without explicitly tracking these. Moreover, \(F^{SCF}\) vary for different climatic zones (\(c\)) specified by the IPCC (see \ref{append:climatemap}). The actual SOC stocks as thus calculated based on a given reference SOC stock by

\begin{equation}
SOC^{\text{target}}_{i} = \sum_{c,m} T_{c,i} \cdot SOC^{\text{ref}} \cdot F^{SCF}_{c,m} \qquad\forall\quad t,
\label{eq:tier1}
\end{equation}

with \(T_{c,i}\) being the translation matrix for grid cells \(i\) into corresponding climate zones \(c\). For this analysis we use the default \(F^{SCF}\) from the Tier 1 method of \citep{ipcc_2006_2006}, and \citep{ipcc_2019_2019} as a comparison and consistency check for our more detailed budget approach.

>>>>

%% 2-column papers and discussion papers
\documentclass[gc, manuscript]{copernicus}

@@

\Author[1]{Kristine}{Karstens}
\Author[1]{Benjamin Leon}{Bodirsky}
\Author[1]{Jan Philipp}{Dietrich}
\Author[2]{Marta}{Dondini}
\Author[1]{Jens}{Heinke}
\Author[2]{Matthias}{Kuhnert}
\Author[1]{Christoph}{Müller}
\Author[1]{Susanne}{Rolinski}
\Author[2]{Pete}{Smith}
\Author[1]{Isabelle}{Weindl}
\Author[1]{Hermann}{Lotze-Campen}
\Author[1]{Alexander}{Popp}

\affil[1]{Potsdam Institute for Climate Impact Research (PIK), Member of the Leibniz Association, P.O. Box 60 12 03, 14412 Potsdam, Germany}
\affil[2]{Institute of Biological \& Environmental Sciences, University of Aberdeen, Aberdeen, UK}
\affil[3]{Humboldt-Universität zu Berlin, Department of Agricultural Economics, Unter den Linden 6, 10099 Berlin, Germany}

@@

\begin{abstract}
Soil organic carbon (SOC) is one of the largest terrestrial carbon stocks on Earth. It is three times larger than the vegetation carbon pool and more than twice as big as the atmospheric carbon pool, when looking into the first meter of the Earth's soil profile. SOC has been depleted by anthropogenic land cover change and agricultural management. However, the latter has so far not been well represented in global carbon stock assessments. While SOC models often simulate detailed biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated at the global scale. Here we develop a spatial explicit data set for agricultural management on cropland, considering crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We combine it with the IPCC Tier 2 steady-state soil model to create a half-degree resolution data set of SOC stocks and SOC stock changes for the first 30 cm of mineral soils. We estimate that due to arable farming, soils have around 26 GtC relative to a counterfactual natural state in 1975. Yet, within the period 1975--2010 this SOC debt has been decreasing again by a net quantity of 4 Gt SOC, which can be mainly traced back to an increased input of C in crop residues due to higher crop productivity. We also find that SOC is very sensitive to management decisions such as residue returning indicating the necessity to incorporate better management data in soil model simulations.
\end{abstract}

@@

Soil Organic Carbon (SOC), the amount of organic carbon stored in the Earth's soil, is the largest terrestrial organic carbon pool. It exceeds the carbon in the atmospheric and vegetation pools multiple times (\citep{batjes_total_1996}). Even small changes in processes affecting SOC lead therefore to substantial shifts in the terrestrial carbon cycle and influence the amount of CO\textsubscript{2} in the atmosphere (\citep{friedlingstein_global_2019}, \citep{minasny_soil_2017}). The specific amount of carbon stored in soils globally is quantified with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile (\citep{batjes_total_1996}, \citep{sanderman_soil_2017}).

Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles to determine SOC variations over space and time. Recent studies have focused on the evaluation of total SOC stocks of the world as well as on the spatial disaggregation of soil properties such as SOC content (\citep{batjes_harmonized_2016}, \citep{hengl_soilgrids250m_2017}, \citep{fao_global_2018}). However, these studies often do not include human interventions, like land cover change and agricultural management, in their analysis. Compared to climatic and geological drivers, they alter terrestrial carbon pools over much shorter time scales and are currently one of most dominant drivers of SOC changes on managed land.

The anthropogenic impact can be measured by the SOC debt (also referred to as SOC component of land-use change emissions), which is the amount of organic carbon soils have lost under cultivation compared to a potential natural vegetated state. \citep{sanderman_soil_2017} identified the anthropogenic SOC debt for the first meter of the soil profile due to land cover change at around 116 GtC (37 GtC for the first 30 cm), compared to previous estimates of 60--130 GtC for the first meter (\citep{lal_world_2001}).

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs) and Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions (\citep{friedlingstein_global_2019}). While providing estimates of the magnitude of SOC losses due to land cover change, most models lack a detailed consideration of agricultural management. Earlier DGVM and ESM based assessments only considered changes in land cover, but ignored the removal of biomass at harvest (\citep{strassmann_simulating_2008}, \citep{betts_climate_2015}). BKMs are designed to estimate LUC realted emissons but ignore changes in SOC due to climate change, CO\textsubscript{2} fertilization, and agricultural management (\citep{friedlingstein_global_2019}, \citep{houghton_carbon_2012}, \citep{hansis_relevance_2015}).

Managed agricultural systems were introduced in greater detail to DGVMs and ESMs to improve the assessment of the terrestrial carbon balance (e.g.~\citep{bondeau_modelling_2007}, \citep{lindeskog_implications_2013}). \citep{pugh_simulated_2015} explicitly consider agricultural management in the form of tillage, irrigation and biomass extraction at harvest, but worked with stylized scenarios rather than with historic management data. They also showed the importance of accounting for the land-use history, as many carbon emissions from agricultural soils are caused by historic LUC and the slow decline of SOC under cropland before it reaches a new equilibrium.

In global-scale carbon cycle assessments, management systems are typically represented as spatially explicit patterns that are static in time (e.g.~growing seasons ({[}portmann\_mirca2000global\_2010{]}) multiple cropping systems (\citep{waha_multiple_2020}), irrigation systems (\citep{jagermeyr_water_2015})) or as stylized scenarios (e.g.~\citep{pugh_simulated_2015}, \citep{lutz_simulating_2019}).

More data sets on spatially explicit agricultural management time series with global coverage become available (e.g.~on tillage systems, see \citep{porwollik_generating_2018}, \citep{prestele_spatially_2018}) and model approaches are increasingly developed to project the dynamics of management systems into the future (e.g.~\citep{iizumi_modeling_2019}, \citep{minoli_modelling_2019}), but have -- to our knowledge -- not found their way into comprehensive assessments of the terrestrial carbon cycle in DGVMs and BKMs.

Field-scale models (\citet{del_grosso_simulated_2001}{]}, \citep{coleman_simulating_1997},\citep{smith_estimating_2010}, \citep{taghizadeh-toosi_c-tool_2014}) are able to better account for historic agricultural management if detailed information on crop yield levels, fertilizer inputs and various other on-farm measures is available for the studied sites. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global domain remains a complex challenge.

The objective of our study is to provide the first global, spatially explicit SOC and SOC debt map that considers spatially explicit and time-variant historical agricultural management. To achieve this objective we create and provide a comprehensive dataset of the global gridded management time series data, including crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We simulate SOC stocks, dynamics and the SOC debt for 1975--2010. Using a scenario analysis, we decompose the contribution of different management activities. Our analysis highlights the importance of considering agricultural management for improving estimates of SOC stocks. We conclude, that historic agricultural management let to increasing SOC stocks in the most recent decades.
\newpage

\hypertarget{methods}{%
\section{Methods}\label{methods}}

In Sect. \ref{sec:carbonbudget} we introduce the basic concept of SOC dynamics as applied in this study and explained in more detail within the refinement of the IPCC guidelines vol.~4 (\citep{calvo_buendia_ipcc_2019}). We additionally describe how we configured and extended the steady-state method (for model code see \citet{mrsoil_2020}). In Sect. \ref{sec:tier1} we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines (\citep{eggleston_hs_ipcc_2006}, \citep{calvo_buendia_ipcc_2019}).
Section \ref{sec:agrimanagement} provides a detailed description of the global, gridded management data used to drive the model, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices (for model code see \citet{mrcommons_2020}). In Sect. \ref{sec:scenarios} we define the scenarios used to complement our historic model results.

\hypertarget{sec:carbonbudget}{%
\subsection{SOC stocks and stock changes following the steady-state method}\label{sec:carbonbudget}}

Following the Tier 2 steady-state approach of the refinement of the IPCC guidelines vol.~4 (\citep{calvo_buendia_ipcc_2019}; referred to as \textit{steady-state method} in the following), we estimate soil organic carbon (\(SOC\)) stocks and their change over time for cropland at half-degree resolution from 1975 to 2010. We restrict our analysis to the first 0-30 cm of the soil profile. Moreover, we assume the current \(SOC\) state converges towards a steady state, which itself depends on biophysical, climatic and agronomic conditions.
Therefore, we take the following three steps for each year of our simulation period:
(1) We calculate annual land-use type-specific steady states and decay rates for \(SOC\) stocks (Sect. \ref{sec:steadystates});
(2) we account for land conversion by transferring \(SOC\) from and natural vegetation (Sect. \ref{sec:carbontransfer}),
(3) we estimate \(SOC\) stocks and changes based on the stocks of the previous time step, the steady state stocks and the decay rate (Sect. \ref{sec:totalsoc}).
To initialize the first year of our simulation period we use a spin-up period of 74 years (Sect. \ref{sec:initsoc}).

\hypertarget{sec:steadystates}{%
\subsubsection{Steady-state SOC stocks and decay rates}\label{sec:steadystates}}

In a simple first order kinetic approach the steady-state soil organic carbon stocks \(SOC^{\mathrm{eq}}\) are given by
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
\label{eq:inoutflow}
\end{equation}
with \(C^{\textrm{in}}\) being the carbon inputs to the soil, \(k\) denotes the soil organic carbon decay rate. This equation is valid for all grid cells \(i\) and all years \(t\). We use the steady-state method for our calculations, which assumes three soil carbon sub-pools \(sub\) (active, slow and passive) and interactions between them, following the approach in the Century model \citep[(][]{parton_analysis_1987}). Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type \(lu\) specific.
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

Carbon inputs for cropland are below and above ground residues left or returned to the field (see Sect. \ref{sec:residues}) and manure inputs (see Sect. \ref{sec:livstmanure}); for natural vegetation litterfall (including fine root turnover) (\citep{schaphoff_lpjml4_2018}) is the only source of carbon inflow to the soil. Following the IPCC guidelines \citep{calvo_buendia_ipcc_2019}, carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content. For each component the sum of all carbon input sources is allocated to the respective \(SOC\) sub-pools via transfer coefficients. This implies that both the amount of carbon and its structural composition determine the effective inflow into the different pools. Data sources for all considered carbon inputs as well as for lignin and nitrogen content are listed in Table \ref{tab:datasourceinputs}.

  \begin{table*}[h]
  \caption{Type and data sources for carbon inputs and parameterization to different land-use types }
  \begin{tabular}{p{0.15\textwidth} p{0.20\textwidth} p{0.25\textwidth} p{0.25\textwidth}}
  \tophline
  \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
  \middlehline
  \multirow{3}{*}{Cropland} & above ground residues & \multirow{3}{*}{\begin{minipage}[t]{0.28\columnwidth}\raggedright
                                                                      \cite{faostat_faostat_2016}, \\
                                                                      \cite{schaphoff_lpjml4_2018} (LPJmL),\\
                                                                      \cite{weindl_livestock_2017} \end{minipage}} & 
                                                      \multirow{2}{*}{\begin{minipage}[t]{0.28\columnwidth}\raggedright
                                                                      lg/c generic values according to Table 5.5B\\
                                                                      in \cite{calvo_buendia_ipcc_2019},\\
                                                                      crop-type specific n/c-ratios from
                                                                      \cite{dietrich_mrcommons_2020}\end{minipage}} \\ \cline{2-2}
                            & below ground residues &  &  \\ \cline{2-2}
                            & manure                &  & Table 5.5C in \cite{calvo_buendia_ipcc_2019} \\ 
                            \hline
  Natural vegetation        & annual litterfall &  \cite{schaphoff_lpjml4_2018} (LPJmL) 
                            & \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut 
                                \cite{calvo_buendia_ipcc_2019} (CENTURY parameterization) \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}

The sub-pool specific decay rates \(k_{sub}\) are influenced by climatic conditions, biophysical and biochemical soil properties as well as management factors that all vary over space (\(i\)) and time (\(t\)). Following the steady-state method \citep{calvo_buendia_ipcc_2019}, we consider temperature (\(temp\)), water (\(wat\)), sand fraction (\(sf\)) and tillage (\(till\)) effects to account for spatial variation of decay rates. Thus, \(k_{sub}\) are given by

\begin{equation}
\begin{aligned}
& k_{i,t,\mathrm{active},lu}  & = &~ k_{\mathrm{active}}  ~ &\cdot~ temp_{i,t} ~ &\cdot~ wat_{i,t,lu} ~ &\cdot~ till_{i,t,lu} ~ & \cdot~ sf_{i}\\
& k_{i,t,\mathrm{slow},lu}    & = &~ k_{\mathrm{slow}}    ~ &\cdot~ temp_{i,t} ~ &\cdot~ wat_{i,t,lu} ~ &\cdot~ till_{i,t,lu} ~ &\\
& k_{i,t,\mathrm{passive},lu} & = &~ k_{\mathrm{passive}} ~ &\cdot~ temp_{i,t} ~ &\cdot~ wat_{i,t,lu} ~ & ~ & 
\label{eq:decayrates}
\end{aligned}.
\end{equation}

For natural vegetation, we assume rainfed and non-tilled conditions, whereas for cropland, we distinguish the effect of different tillage (see Sect. \ref{sec:tillage}) and irrigation (see Sect. \ref{sec:irrigation}) practices on decay rates. We calculated area weighted means for \(till\) and \(wat\) on cropland for each grid cell, using area shares for the different tillage and irrigation pratices. Data sources as well as used parameters for the different decay drivers for all land-use types are listed in Table \ref{tab:datasourcedecay}; equations are displayed by equation 5.0B--5.0F in \citep{calvo_buendia_ipcc_2019}.

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types}
 \begin{tabular}{l l l l}
 \tophline
  land-use types   & type of decay driver & parameter use to represent driver & data source \\
 \middlehline
 \multirow{2}{*}{all} & Soil quality & Sand fraction of the first 0-30 cm 
                                     & \cite{hengl_soilgrids250m:_2017} \\
                      \cline{2-4}

                      & Mircobial activity & air temperature & \cite{harris_version_2020} \\
                      \cline{2-4}

                      & Water restriction & precipitation \& potential evapotranspiration & \cite{harris_version_2020} \\
                      \cline{1-4}
\multirow{2}{*}{\begin{minipage}[t]{0.2\columnwidth}\raggedright\strut Cropland\\(additionally)\strut\end{minipage}} & Water restriction*  & irrigation  & Sect. \ref{sec:irrigation} \\ 
                      \cline{2-4}

                      & Soil disturbance & tillage & Sect. \ref{sec:irrigation} \\
 \bottomhline
 \end{tabular}
 \belowtable{}
 \label{tab:datasourcedecay}
 \end{table*}

\hypertarget{sec:carbontransfer}{%
\subsubsection{SOC transfer between land-use types}\label{sec:carbontransfer}}

We calculate \(SOC\) stocks based on the area shares of land-use types (\(lu\)) within our grid cells (\(i\)). If land is converted from one land-use type \(lu=\{crop,natveg\}\) into the other \(!lu=\{natveg,crop\}\), a respective share of the \(SOC\) is reallocated. We account for land conversion at the beginning of each time step \(t\) by calculating a preliminary stock \(SOC_{t^*}\) via

\begin{equation}
SOC_{i,t^*,sub,lu} = SOC_{i,t-1,sub,lu} - \frac{SOC_{i,t-1,sub,lu}}{A_{i,t-1,lu}} \cdot  AR_{i,t,lu} + \frac{SOC_{i,t-1,sub,!lu}}{A_{i,t-1,!lu}} \cdot  AE_{i,t,lu}
\label{eq:ctransfer}
\end{equation}

with \(A_{lu}\) being the land-use type specific areas, \(AR_{lu}\) and \(AE_{lu}\) the area reduction resp. area expansion of the two land-use types. Data sources and methodology on land-use states and changes are described in Sect. \ref{sec:landuse}.

\hypertarget{sec:totalsoc}{%
\subsubsection{Total SOC stocks and stock changes}\label{sec:totalsoc}}

\(SOC\) converge towards the calculated steady-state stock \(SOC^{\mathrm{eq}}\) for each grid cell \(i\), each annual time step \(t\), each land-use type \(lu\) and each sub-pool \(sub\) like

\begin{equation}
SOC_{i,t,sub,lu} = SOC_{i,t-1,sub,lu} + (SOC^{\mathrm{eq}}_{i,t,sub,lu} - SOC_{i,t-1,sub,lu}) \cdot k_{i,t,sub,lu} \cdot 1\unit{a}.
\label{eq:SOCstate}
\end{equation}

Note that the decay rates have to be multiplied by one year (\(1a\)) to form a dimensionless factor.
Reformulating this equation, we obtain a mass balance equation as follows

\begin{equation}
SOC_{i,t,sub,lu} = SOC_{i,t-1,sub,lu} - \underbrace{SOC_{i,t-1,sub,lu} \cdot k_{i,t,sub,lu} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{\mathrm{eq}}_{i,t,sub,lu} \cdot k_{i,t,sub,lu} \cdot 1\unit{a}}^{\text{input (using equation (1))}}.
\label{eq:steadystate2budget}
\end{equation}

The global \(SOC\) stock for each time step \(t\) can then be calculated via

\begin{equation}
SOC_{t} = \sum_{i} \underbrace{\sum_{lu} \overbrace{\sum_{sub} SOC_{i,t,sub,lu}}^{\text{$SOC_{i,t,lu}$ -- land-use type specific SOC stock within cell}}}_{\text{$SOC_{i,t}$ -- total SOC stock within cell}}.
\label{eq:totalstock}
\end{equation}
